% COMP-767 final paper — ICML 2026 format
% Requires: icml2026.sty, icml2026.bst, fancyhdr.sty (from official ICML author kit in this directory)
% Compile: pdflatex -> bibtex -> pdflatex x2

\documentclass{article}

\usepackage{microtype}
\usepackage{graphicx}
% \usepackage{subcaption} % ICML recommends for subfigures; install caption bundle if needed
\usepackage{booktabs}

\usepackage{hyperref}

\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Blind submission (default). For preprint: \usepackage[preprint]{icml2026}
% Camera-ready: \usepackage[accepted]{icml2026}
\usepackage{icml2026}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
% Optional (ICML example): \usepackage{mathtools}
% Load after hyperref/amsmath: \usepackage[capitalize,noabbrev]{cleveref}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\icmltitlerunning{Linguistic MDL Explanations via Occam-Style Prompt Discovery}

\begin{document}

\twocolumn[
  \icmltitle{Linguistic MDL Explanations with LLMs via Occam-Style Prompt Discovery}

  \begin{icmlauthorlist}
    \icmlauthor{Yassine Meliani}{mcgill}
    \icmlauthor{George Kontorousis}{mcgill}
    \icmlauthor{Eric Elmoznino}{mila}
  \end{icmlauthorlist}

  \icmlaffiliation{mcgill}{Department of Computer Science, McGill University, Montreal, QC, Canada}
  \icmlaffiliation{mila}{Mila Quebec, Montreal, QC, Canada}

  \icmlcorrespondingauthor{Yassine Meliani}{yassine.meliani@mail.mcgill.ca}
  \icmlcorrespondingauthor{George Kontorousis}{george.kontorousis@mail.mcgill.ca}
  \icmlcorrespondingauthor{Eric Elmoznino}{eric.elmoznino@mila.quebec}

  \icmlkeywords{Minimum description length, large language models, interpretability, reinforcement learning, prompt discovery}

  \vskip 0.3in
]

\printAffiliationsAndNotice{}

\begin{abstract}
  % TODO: Revise after experiments — keep aligned with final results.
  Prompted large language models (LLMs) behave like universal function approximators: the prompt acts as natural-language ``parameters'' governing conditional predictions.
  We frame linguistic explanations as hypotheses about data and apply the Minimum Description Length (MDL) principle using frozen LLM likelihoods.
  A trainable policy $p_\theta(z \mid x)$ proposes an explanation $z$ for dataset $x$; a frozen evaluator $p^\star$ scores $z$ with a two-part code: explanation complexity $-\log p^\star(z)$ and data complexity $-\log p^\star(x \mid z)$ (with token masking for prediction targets).
  We optimize $p_\theta$ with reinforcement learning to discover concise, predictive explanations and compare to in-context learning and Monte Carlo hypothesis search.
  % TODO: Summarize key empirical findings (compression, generalization, interpretability).
\end{abstract}

\section{Introduction}
\label{sec:intro}

% Motivation: prompts as models; heuristic explanation vs.\ principled simplicity + fit.
% Occam / MDL; Bayesian view as maximizing $p^\star(z \mid x)$.
% High-level method: policy $p_\theta$, frozen $p^\star$, RL on MDL reward.
% Contributions (enumerate explicitly).

\section{Related Work}
\label{sec:related}

See, e.g., \citet{petrov2024} on prompt-conditioned behavior; RL approaches to prompts \citep{hu2023}; Monte Carlo hypothesis search in textual environments \citep{levy2025}; and dataset-centric automated discovery \citep{autodiscovery2025}.
% Expand: MDL / description length; other explanation / interpretability lines.

\section{Method}
\label{sec:method}

\subsection{Problem setting}
\label{sec:problem}

% Dataset $x$, mask $m$ over tokens for likelihood of targets (labels, next observation, etc.).

\subsection{MDL objective and reward}
\label{sec:mdl}

Let $x$ denote a dataset (rendered as text), $z$ a proposed linguistic explanation, and $m$ a mask over tokens that count toward the predictive likelihood (e.g., labels or next-observation fields).
With a frozen evaluator LM $p^\star$, we define the reward
\begin{equation}
  R(z, x, m; p^\star) = -\log p^\star(z) - \log p^\star(x \mid z),
  \label{eq:mdl-reward}
\end{equation}
The first term penalizes explanation complexity under $p^\star$; the second penalizes poor prediction of $x$ given $z$, evaluated only on tokens selected by the mask $m$.
% TODO: Clarify exact factorization / conditional used in implementation (which spans are conditioned on).

\subsection{Explanation policy and training}
\label{sec:policy}

% Policy $p_\theta(z \mid x)$: architecture, fine-tuning base model, prompt template for embedding $x$.
% RL algorithm (e.g., PPO / REINFORCE), variance reduction, baselines.
% Why frozen $p^\star$ mitigates reward hacking.

\section{Experiments}
\label{sec:experiments}

\subsection{Datasets}
\label{sec:datasets}

\paragraph{Toy textual environment.}
% Levy-style textual observations / transitions; mask on ``next observation'' (or equivalent) span.

\paragraph{Natural-language or IO-pair benchmarks.}
% Subset of AutoDiscovery-style datasets; mask on output span for supervised-style formatting.

\paragraph{Synthetic rules (optional).}
% Parity, modular arithmetic, string transforms — sanity checks for rule recovery.

\subsection{Baselines}
\label{sec:baselines}

\begin{enumerate}
  \item \textbf{RL-MDL (ours):} policy $p_\theta(z \mid x)$ trained to maximize~\eqref{eq:mdl-reward}.
  \item \textbf{In-context learning (ICL):} no explicit $z$; predictions from training examples in context.
  \item \textbf{Monte Carlo search:} sample or search candidate $z$; select by same MDL score without learned policy.
\end{enumerate}

\subsection{Metrics}
\label{sec:metrics}

\paragraph{Training-set compression.}
For explanation-based methods (ours and search),
\begin{equation}
  -\log p^\star(z) - \sum_{i=1}^{N} \log p^\star(x_i \mid z).
  \label{eq:compression-two-part}
\end{equation}
For ICL, an online one-part code,
\begin{equation}
  -\log p^\star(x_1) - \sum_{i=1}^{N-1} \log p^\star(x_{i+1} \mid x_{1:i}).
  \label{eq:compression-icl}
\end{equation}

\paragraph{Generalization.}
For held-out $x' = (x'_1, \ldots, x'_M)$, explanation-based methods report
\begin{equation}
  \sum_{i=1}^{M} -\log p^\star(x'_i \mid z),
\end{equation}
and ICL reports $\sum_{i=1}^{M} -\log p^\star(x'_i \mid x)$.

\paragraph{Interpretability proxies.}
% Token length $|z|$; optional human study.

\subsection{Implementation details}
\label{sec:impl}

% Models ($p^\theta$, $p^\star$), hyperparameters, compute, seeds, training steps.

\section{Results}
\label{sec:results}

% Tables/figures: compression, generalization, $|z|$.
% Ablations if any (mask design, reward components, search budget vs.\ policy).

\section{Discussion and limitations}
\label{sec:discussion}

% Failure modes, scalability, dependence on $p^\star$, ethical / misuse notes if relevant.

\section{Conclusion}
\label{sec:conclusion}

% Restate contributions and outcomes; brief future work.

\section*{Accessibility}
Authors will follow ICML accessibility guidance for the camera-ready version as applicable.

\section*{Software and Data}
If accepted, we plan to release code and data references via anonymized supplementary material during review where appropriate, and public URLs in the camera-ready version.

% Acknowledgements only in accepted / camera-ready; text suppressed under blind review by ICML style.
\section*{Acknowledgments}
\textbf{Do not} include acknowledgements in the initial version submitted for blind review.
% TODO: Funding, compute, colleagues (camera-ready).

\section*{Impact Statement}
This paper presents work whose goal is to advance the field of Machine Learning.
There are many potential societal consequences of automated explanation discovery with LLMs, including misuse for generating plausible but incorrect rationales; we encourage careful human oversight in high-stakes settings and will expand this discussion in the camera-ready version as appropriate.

\bibliography{references}
\bibliographystyle{icml2026}

\newpage
\appendix
\onecolumn

\section{Additional experimental details}
\label{app:details}

% Extended hyperparameters, prompts, dataset formatting examples.

\section{Extended results}
\label{app:results}

% Extra plots, per-dataset breakdowns, qualitative explanation samples.

\end{document}
